---
title: "Project_2_Group"
author: "Priya Bhat"
date: "2025-12-01"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

# LARs using iris dataset

## Libraries and packages
```{r}
#install.packages("lars")
library(lars)
data(iris)

```



## Assigning values
```{r}
y <- iris$Sepal.Length

x <- as.matrix(iris[, c("Sepal.Width", "Petal.Length", "Petal.Width")])

dim(x)
head(x)
```


## LARS
```{r}
lars_model <- lars(x, y, type = "lar")
lars_model

#plot
plot(lars_model)
legend("topleft", legend = colnames(x), col = 1:3, lty = 1)
```


## Choosing the best model
```{r}
set.seed(123)

cv_model <- cv.lars(x, y, type = "lar")

plot(cv_model$index, cv_model$cv, type = "b",
     xlab = "Fraction of L1 norm",
     ylab = "Cross Validation Error",
     main = "Cross Validation for LARS")

```
