---
title: "Project_2_Group"
author: "Priya Bhat"
date: "2025-12-01"
#output: beamer_presentation
fontsize: 9pt
classoption: "aspectratio=169"
output:
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r echo = FALSE}
#install.packages("lars")
library(lars)
data(iris)

```


## LARS using iris Dataset

LARS requires us to have:

- a numeric vector for the response (y)
- a matrix of predictor variables (x)

```{r echo = TRUE}
y <- iris$Sepal.Length
x <- as.matrix(iris[, c("Sepal.Width", "Petal.Length", "Petal.Width")])

```


```{r echo = FALSE}
lars_model <- lars(x, y, type = "lar")
```

## Model Summary
```{r}
lars_model
```


## iris LARS Plot

We can see that **Petal.Length** has the strongest correlation with Sepal.Length. **Sepal.Width** has the 2nd strongest correlation, and **Petal.Width** is the least correlated.
```{r}
plot(lars_model)
legend("topleft", legend = colnames(x), col = 1:3, lty = 1)
```

```{r echo = FALSE}
data(mtcars)

head(mtcars)


```


```{r echo = FALSE}
y_val <- mtcars$mpg
x_val <- as.matrix(mtcars[, -1])

#lars model
lars_model_mt <- lars(x_val, y_val, type = "lar")

```

## LARS using mtcars Dataset

In this model, we are looking to predict fuel efficiency:

- Miles per Gallon (mpg) as our response variable (y)
- Matrix of all predictor variables (x)
```{r}
head(mtcars)
```

## Model Summary

```{r}
summary(lars_model_mt)
```


## mtcars LARS Plot

Here we see that **wt** and **cyl** enter the model first, meaning they are more strongly correlated to **mpg** than other variables such as **disp** and **qsec**. 
```{r}
plot(lars_model_mt)
legend("bottomleft", legend = colnames(x_val), col = 1:10, lty = 1, pch = 16, cex = 0.70)
```

## Key Findings

- Through these two examples, we can see that LARS makes it very easy to identify which variables have the most influence over our response variable.

- It shows us how important each predictor is, while also showing us where each variable enters model.

- LARS also adds variables into the model in a balanced way without over fitting.

